Synthesizing Quotient Lenses

### 2017-10-05 papers/QuotientLensSynth/main.tex
Did more explaining for QREs. Added figures for QRE semantics.

\section{Introduction}
\saz{We need a more gentle ramp-up to the example here.}

\begin{itemize}
\item Lens programming languages [e.g. Boomerang] are designed to make it possible to express
  bidirectional data transformations using a single program.

\item Lenses have found practical application in Augeas \saz{and ...}

\item The source and target types of a lens express constraints on the format of
  the data.

\item Lens programming is still quite tricky, because lens composition operators
  are subject to side conditions (e.g. unique decomposition) that are not so
  easy to reason about.

\item Our prior work~\cite{popl18} has shown how to synthesize lenses for the
  \textit{bijective} case.

\item This still isn't as expressive as one would want in practice, because many
  lens transformations aren't truly bijective.

\item There are at least two challenges with programming quotient lenses: (1)
  defining the quotient languages (which serve as the types of the lenses)
  themselves is not so easy, and (2) the lenses themselves are still not easy to
  write by hand.

\end{itemize}

String quotient lenses enable one to express bidirectional string
transformations that obey the lens laws up to equivalence relations. However,
the various approaches that have been proposed to express quotient lenses
make it difficult for the programmer to express complicated equivalence
relations over the data.

For example, when converting \textsc{Bib}\TeX{} files to EndNote files, the data
may contain arbitrary amounts of whitespace characters in between the actual
data. In addition, the fields may be ordered in an arbitrary order. For
instance, we may wish to convert this \textsc{Bib}\TeX{} entry: \saz{I propose
  we display such examples in verbatim:}

@@

to this EndNote entry:

@@

With the existing tools, it is cumbersome to express this, and other complicated
equivalence relations.  \saz{It might be a good idea to show what this would
  look like in all its gory detail, just to make the problem very
  concrete. Perhaps in a figure?}

In this paper, we introduce the formalism of Quotient Regular Expressions (or
QREs) to solve this problem. QREs enable the programmer to express an
equivalence relation at internal terms of a regular expression, thus giving
the programmer more control over where quotienting occurs in the data.\\
For example, the fields of a \textsc{Bib}\TeX{} article like in the example
above might be expressed by the following QRE:

$$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER}, \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)},$$

\noindent where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name. The QRE of an EndNote article is similarly expressed, but with the
regular expressions for the respective fields substituted with the appropriate
EndNote equivalent.  \saz{Do we really mean all \re{WSP} maps to newline here?
  What about the spaces in the title, etc.?  Would this be confusing without
  more explanation?}

\sam{[Mention and give examples of quotient lenses here?]}

\subsection{Contributions}
\begin{enumerate}
  \item
  We introduce the language of Quotient Regular Expressions (QREs) which enables
  us to express a set of equivalence relations on regular languages using a
  compact, easy-to-use syntax.
  \item
  We describe a class $\mathcal{Q}$ of quotient lenses whose types are given by
  QREs. We derive a normal form for quotient lenses in $\mathcal{Q}$, and
  demonstrate that elements of $\mathcal{Q}$ that are in this normal form are
  closed under left quotienting, right quotienting, composition and regular
  operators.
  \item
  Using the normal form for quotient lenses in $\mathcal{Q}$, we reduce the
  synthesis of quotient lenses in $\mathcal{Q}$ to the synthesis of bijective
  lenses between regular languages. 
  \item
  We describe an implementation of QREs, the language $\mathcal{Q}$ of quotient
  lenses, and an algorithm for synthesizing quotient lenses in $\mathcal{Q}$ in
  the Boomerang language.
  \item
  We assess the usabibility of the QRE synthesis via a case study in which we
  selected several data formats used by data.gov and synthesized quotient lenses
  between them.
\end{enumerate}

\section{Quotient Regular Expressions}

\saz{Need a transition here:  This section introduces QREs.  They are regular
  expressions augmented with syntax that lets them simultaneously denote
  quotient languages.  
   }

  \paragraph{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep q \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{(q_1, \ldots, q_n)}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions whose
domain and codomain are regular languages, and $s$ ranges over character
strings.

\saz{  Every regular expression $R$ is a QRE.  The main new syntactic forms are: $q \mapsto s$, which
}

@@

Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely, each
Quotient Regular Expression (or QRE) $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $Canonizer(q)$.\\
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
The transformations that will be performed on the data will require
that the data is matched against regular expressions. Consequently, we require
that regular expressions used by QREs be \textit{strongly unambiguous}, a
condition which ensures that if a string $s$ matches a regular expression $R$,
then $s$ matches $R$ uniquely.

\saz{Solomon: you should use blank lines instead of backslashes to separate
  paragraphs so latex can format them properly.}

@@

We now give the semantics of QREs as well as inference rules for deriving them:
\textcolor{red}{[Informal Meanings/Example Usages Needed]}

\begin{figure}[t]
  \centering
  \[
    \begin{array}{r@{\quad}c@{\quad}cc}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      q \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q_1^* & W(q_1)^* & K(q_1)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions.}
  \label{fig:wk}
\end{figure}

\begin{enumerate}
  \item $q = R$
\begin{center}
\begin{enumerate}
  \item $W(R) = R$,
  \item $K(R) = R$, 
  \item
For all $w, w' \in W(R)$, $w \; EqRel(q) \; w'$ if and only if $w =
w'$.
\item
$Canonizer(R)(r) = r$
\end{enumerate}
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}
\end{center}
\item $q=q \mapsto s$
\begin{center}
\begin{enumerate}
  \item $W(q \mapsto s) = W(q)$,
  \item $K(q \mapsto s) = s$, 
  \item
For all $w, w' \in W(q \mapsto S)$, $w \; EqRel(q \mapsto S) \; w'$,
\item
$Canonizer(q \mapsto S)(r) = s$
\end{enumerate}
\begin{prooftree}
\AxiomC{$s \in W(q)$}
\UnaryInfC{$q \mapsto s$ is well formed}
\end{prooftree}
\end{center}
\item $q=squash(R, R', f)$
\begin{center}
\begin{enumerate}
  \item $W(squash(R, R', f)) = R \sep R'$,
  \item $K(squash(R, R', f)) = R'$, 
  \item
For all $w, w' \in W(squash(R, R', f))$, $w \; EqRel(q) \; w'$ if and only
if $f(w) = w'$ or $w = w'$
\item
$Canonizer(squash(R, R', f))(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', f)$ is well formed}
\end{prooftree}
\end{center}
\item $q=perm \; (q_1, \ldots, q_n) \; with \; q$. Given regular expressions
$R_1, \ldots, R_n$ and $S$, let $\prod({R_i}, S) = R_1 \cdot S \cdot \ldots
\cdot S \cdot R_n$.
\begin{center}
\begin{enumerate}
  \item $W(perm \; (q_1, \ldots, q_n) \; with \; q) = \bigcup \limits_{\sigma
  \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot
  W(q_{\sigma(n)}),$
  \item $K(perm \; (q_1, \ldots, q_n) \; with \; q) = K(q_1) \cdot K(q_1)
  \cdot \ldots \cdot \cdot K(q) \cdot K(q_n)$, 
  \item
For all $w, w' \in W(q)$,  $w \; EqRel(q) \; w'$ if and only if
   $$w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot
    r_{\sigma(n)} \text{ and } r' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots
    \cdot s'_{n-1} \cdot {r'}_{\theta(n)}$$ for some $\sigma, \theta \in
    S_n$, with $r_i \; EqRel(q_i) \; r'_i$ and $s_k \; EqRel(q) \; s'_{k}$.
\item
$Canonize(q)(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonize(q_1)(r_1)) \cdot (Canonize(q)(s_1)) \cdot \ldots \cdot \cdot \cdot
(Canonize(q_n)(r_n))$
\end{enumerate}
\begin{prooftree}
\AxiomC{$\forall \sigma \neq \theta, \; \prod(\{W(q_{\sigma(i)})\}, W(q))\cap
 \prod(\{W(q_{\theta(i)})\}, W(q)) = \varnothing$}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\BinaryInfC{$permute \; (q_1, \ldots, q_n) \; with \;  q$ is well formed}
\end{prooftree}
\end{center}

\item $q = normalize(R, R', f)$
\begin{center}
$W(normalize(f)) = R$
$K(normalize(f)) = R'$
\item For all $w, w' \in W(q)$, $w \; EqRel(q) \; w'$ if and only if
$f(w)=f(w')$
$Canonize(normalize(f)) = f$
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$normalize \; (R,R', f)$ is well formed}
\end{center}
\item $q=q' \circ q$
\begin{center}
  \item $W(q' \circ q) = W(q)$,
  \item $K(q' \circ q) = K(q')$, 
For all $w, w' \in W(squash(R, R', f))$, $w \; EqRel(q' \circ q) \; w'$ if and
only $$\text{there exist }k, k' \in
  \mathcal{L}(K(q)) \text{ such that } w \; EqRel(q) \; k, \; w' \;
  EqRel(q) \; k', \text{ and } k \; EqRel(q') \; k' $$
$Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)
$
\AxiomC{$q'$ is well formed}
\AxiomC{$K(q) = W(q')$}
\TrinaryInfC{$q' \circ q$ is well formed}
\end{center}
\item $q=q \cdot q'$
\begin{center}
  \item $W(q \cdot q') = W(q) \cdot W(q')$,
  \item $K(q \cdot q') = K(q) \cdot K(q')$, 
For all $w, w' \in W(q \cdot q')$, $w \; EqRel(q' \cdot q) \; w'$ if and
only $$w = r_1
      \cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; EqRel(q) \;
      {r'}_1, \; r_2 \; EqRel(q') \; {r'}_2$$
$Canonizer(q \cdot q') = Canonizer(q) \cdot Canonizer(q')
$
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cdot^! W(q')$}
\AxiomC{$K(q) \cdot^! K(q')$}
\QuaternaryInfC{$q \cdot q$ is well formed}
\end{center}
\item $q=q \sep  q'$
\begin{center}
  \item $W(q \cdot q') = W(q) \sep W(q')$,
  \item $K(q \cdot q') = K(q) \sep K(q')$, 
For all $w, w' \in W(q \sep q')$, $w \; EqRel(q' \sep q) \; w'$ if and
only if 
$$w \; EqRel(q) \; w' \text{ or } \; w \; EqRel(q') \; w'$$
$Canonizer(q \sep q') = Canonizer(q) \sep Canonizer(q')
$
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cap W(q') = \varnothing$}
\TrinaryInfC{$q \cdot q$ is well formed}
\end{center}
\item $q=q'^*$
\begin{center}
  \item $W(q'^*) = W(q')^*$,
  \item $K(q'^*) = K(q')^*$, 
For all $w, w' \in W(q'^*)$, $w \; EqRel(q'^*) \; w'$ if and
only if 
$$w = r_1 \cdot \ldots \cdot r_n, \; w' = {r'}_1 \cdot \ldots
      \cdot {r'}_n \text{ and } r_i \; EqRel(q) \; {r'}_i
      \text{ for all } 1 \leq i \leq n$$
$Canonizer(q'^*) = Canonizer(q')^* $

@@

\end{center}
\begin{claim}

@@

@@

\end{claim}

>>>>

\section{Introduction}
Lense programming languages such as Boomerang enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
program. For instance, lenses enable a programmer to write a program which
enables a doctor to acquire a summary of a patient's data, and also allows the
doctor to reflect updates made to the patient's record back into the main
database.

Lenses therefore enable the  programmer to kill two birds in one
stone by allowing the programmer to express two different data transformations
in one program. Also, when lenses are appropriately constrained, they offer
guarantees about how the two transformations interact with each other. Lenses
have therefore been used effectively as a solution to the \textit{view update
problem}, a classical problem in the database community in which the
programmer wishes to write a program which provides a \textit{view} of data
derived from a \textit{source} to a client, as well as a program which enables
the client to reflect changes made in the view back into the source without
corrupting the source.

However, the constraints that are imposed on lenses are often too strict.
For instance, the programmer may wish to ignore updates to the data that only
change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses} as a
solution to this problem. Quotient lenses enable a programmer to write lenses
which transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

This paper addresses two main challenges that arise when programming quotient
lenses. The first challenge is tied directly to the programming of quotient
lenses in Boomerang, a string lens programming language where the source and
view types are given by regular expressions. While Boomerang enables the
programmer to express a rich set of quotient lenses, the syntax for writing
these lenses is often too verbose and cumbersome. This makes writing quotient
lenses in Boomerang rather difficult.

For example, suppose that the programmer wishes to convert citation records
stored in \textsc{Bib}\TeX{} format to a representation of the same data in the
EndNote format. The programmer wishes to ``quotient'' away unnecessary
whitespace characters in between the various fields in each record, while also
ignoring the order in which these fields occur in the recore. For instance, the
quotient lens which the programmer wishes to derive should convert this
\textsc{Bib}\TeX{} record:

@@

to this EndNote record

@@

While Boomerang enables the programmer to write a lens which exhibits this
behaviour, the syntax tree for the program is large and complicated.

\saz{It might be a good idea to show what this would look like in all its gory
detail, just to make the problem very concrete. Perhaps in a figure?}

In this paper, we introduce the formalism of Quotient Regular Expressions (or
QREs) to address this issue. QREs enable the programmer to express
a rich set of equivalence relations on regular expressions at the level of
syntax. QREs therefore enable the programmer to express more equivalence
relations with less effort. Also, since QREs are defined at the level of
syntax on regular expressions, QREs give the programmer more control over where
the quotienting occurs.

The second challenge that we address in this paper is the synthesis of
\textit{bijective} quotient lenses. These are quotient lenses which match the
equivalance classes of the source data to the equivalence classes of the view
data bijectively as in the \textsc{Bib}\TeX{} to EndNote example above. 
Building on our prior work~\cite{popl18} where we showed how to synthesize
(ordinary?) bijective lenses, we show how to synthesize bijective quotient
lenses given a synthesis task described by a pair of QREs and a set of
examples. To our knowledge this is the first time that this has been done.

Our plan for this paper is as follows:
\begin{enumerate}
  \item
  We introduce the language of Quotient Regular Expressions (QREs) which enables
  us to express a set of equivalence relations on regular languages using a
  compact, easy-to-use syntax.
  \item
  We describe a class $\mathcal{Q}$ of quotient lenses whose types are given by
  QREs. We derive a normal form for quotient lenses in $\mathcal{Q}$, and
  demonstrate that elements of $\mathcal{Q}$ that are in this normal form are
  closed under left quotienting, right quotienting, composition and regular
  operators.
  \item
  Using the normal form for quotient lenses in $\mathcal{Q}$, we reduce the
  synthesis of quotient lenses in $\mathcal{Q}$ to the synthesis of bijective
  lenses between regular languages. 
  \item
  We describe an implementation of QREs, the language $\mathcal{Q}$ of quotient
  lenses, and an algorithm for synthesizing quotient lenses in $\mathcal{Q}$ in
  the Boomerang language.
  \item
  We assess the usabibility of the QRE synthesis via a case study in which we
  selected several data formats used by data.gov and synthesized quotient lenses
  between them.
\end{enumerate}

\section{Quotient Regular Expressions}
This section introduces Quotient Regular Expressions or QREs. QREs are regular
expressions augmented with syntax that lets them simultaneously express an
equivalence relation on the language described by the regular expression.

\sam{More explanation here?}

  \subsection{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep R \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $Canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $Canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $Canonizer(q)$ computes $Canonizer(q)(w)$ as the unique $k$
  in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $Canonizer(q)$.
  \subsection{Meaning and Example Usages for QREs}
  \begin{enumerate}
    \item 
    Every regular expression is a QRE. When a regular expression $R$ is used as
    a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by the equality
    equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. 
    $$FIRST\_NAME \cdot (WHITESPACE^* \mapsto `` \quad ") \cdot LAST\_NAME$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$FIRST\_NAME \cdot `` \quad " \cdot LAST\_NAME$$
    while the second representation may be
    $$FIRST\_NAME \cdot ``," \cdot LAST\_NAME$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WSP}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $WSP$ is a regular expression describing whitespace characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the data should match the whole language $W(q)$ and
the kernel language $K(q)$ unambiguosly so as to make transformations
deterministic.
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

\saz{Every regular expression $R$ is a QRE.  The main new syntactic
forms are:
$R \mapsto s$, which }

@@

\iffalse
Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely,
  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $Canonizer(q)$.
  \fi

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
As we mentioned earlier, when applying the regular constructs $\cdot
\; | \;  *$ to QREs, we require that the regular languages
described by QREs match the data in only one way so as to make transformations
deterministic. Consequently, we require that regular expressions used
by QREs be \textit{strongly unambiguous}, a condition which ensures that if a
string $s$ matches a regular expression $R$, then $s$ matches $R$ uniquely.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $EqRel(q)$ and the
canonizing function $Canonizer(q)$ are given in Figures ~\ref{fig:wk},
~\ref{fig:relations} and ~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{r@{\quad}c@{\quad}cc}

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{(q_1, \ldots, q_n)}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot \cdot W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{(q_1, \ldots, q_n)}{q} ) = &
 K(q_1) \cdot K(q_1) \cdot \ldots \cdot \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions.}
  \label{fig:wk}
\end{figure}

\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & EqRel(q)  \\ \hline
      R & w \; \equiv_R \; w' \iff w = w \\
      R \mapsto s & w \; \equiv_{R \mapsto s} \; w' \text{ for all }w, w'\\
      \squash{R}{R'}{f} & w \; \equiv_{\squash{R}{R'}{f}} \; w' \iff f(w) = w'
      \text{ or } w = w' \\
      \normalize{R}{R'}{f} & w \; \equiv_{\normalize{R}{R'}{f}} \; w' \iff
      f(w)=f(w') \text{ or }w = w'\\
      q_2 \circ  q_1 &  w \; \equiv_{q_2 \circ q_1} \; w' \iff \exists k, k' \in
  \mathcal{L}(K(q_2)) \text{ such that } w \; \equiv_{q_1} \; k, \; w' \;
  \equiv_{q_1} \; k', \text{ and } k \; \equiv_{q_2} \; k'\\
      q_1 \cdot q_2 &  w \; \equiv_{q_1 \cdot q_2} \; w'  \iff w = r_1
      \cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{q_1}
      \; {r'}_1, \; r_2 \; \equiv{q_n} \; {r'}_2\\
      q_1 \sep q_2 &  w \; \equiv_{q' \sep q} \; w' \iff w \; \equiv_{q_1} \; w'
      \text{ or } \; w \; \equiv_{q_2} \; w'\\
      q^* &  w \; \equiv_{q*} \; w' \iff w = r_1 \cdot \ldots \cdot r_n, \; w'
      = {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{q} \; {r'}_i
      \text{ for all } 1 \leq i \leq n\\
    \end{array}
    \]
    \end{center}
    $w \; \equiv_{\perm{(q_1, \ldots, q_n)}{q}} \; w' $ if and only if

    $w = r_{\sigma(1)} \cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}
    \text{ and } w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
    \cdot {r'}_{\theta(n)}$ for some $\sigma, \theta \in S_n$, with $r_i \;
    \equiv_{q_i} \; r'_i$ and $s_k \; \equiv_{q} \; s'_{k}$.
  \caption{QRE Equivalence Relations}
  \label{fig:relations}
\end{figure}
\begin{figure}[t]
  \begin{center}
\[
    \begin{array}{l@{\quad}l} 
      q & Canonizer(q)  \\ \hline
      R & Canonizer(R) = id_{\mathcal{L}(R)} \\
      R \mapsto s & Canonizer(R \mapsto s)(w) = s\\
      \squash{R}{R'}{f} & Canonizer(\squash{R}{R'}{f})(w) = 
\begin{cases}
f(w) & \text{if } w \in \mathcal{L}(R)\\
w & \text{otherwise}
\end{cases}\\
      \normalize{R}{R'}{f} & Canonizer(\normalize{R}{R'}{f}) = f\\
      \perm{(q_1, \ldots, q_n)}{q} & \\
      q_2 \circ  q_1 & Canonizer(q' \circ q) = Canonizer(q') \circ Canonizer(q)\\
      q_1 \cdot q_2 & Canonizer(q' \cdot q) = Canonizer(q') \cdot Canonizer(q)\\
      q_1 \sep q_2 &  Canonizer(q_1 \sep q_2)(w) = 
\begin{cases}
Canonizer(q_1)(w) & \text{if } w \in \mathcal{L}(W(q_1))\\
Canonizer(q_2)(w) & \text{if } w \in \mathcal{L}(W(q_2))\\
\end{cases}\\
      q^* &  Canonizer(q^*) = Canonizer(q)^* \\
    \end{array}
    \]
    \end{center}
    $Canonizer(\perm{(q_1, \ldots, q_n)}{q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (Canonizer(q_1)(r_1)) \cdot (Canonizer(q)(s_1)) \cdot \ldots \cdot
(Canonizer(q_n)(r_n))$
  \caption{QRE Canonizers}
  \label{fig:canonizers}
\end{figure}
\begin{figure}[t]
\centering
\begin{prooftree}
\AxiomC{}
\UnaryInfC{$R$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$s \in W(q)$}
\UnaryInfC{$R \mapsto s$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', f)$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(q_{\sigma(1)}) \cdot W(q)
\cdot \ldots \cdot W(q_{\sigma(n)})\\ \cap W(q_{\theta(1)}) \cdot W(q)
\cdot \ldots \cdot W(q_{\theta(n)}) =\varnothing}}$}
\AxiomC{$q_i, q$ are well formed}
\AxiomC{$\forall \sigma, \; K(q_{\sigma(1)}) \cdot^! K(q) \cdot^! \ldots 
\cdot^! K(q_{\sigma(n)})$}
\TrinaryInfC{$\perm{q_1, \ldots, q_n}{q}$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$normalize \; (R,R', f)$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$K(q) = W(q')$}
\TrinaryInfC{$q' \circ q$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cdot^! W(q')$}
\AxiomC{$K(q) \cdot^! K(q')$}
\QuaternaryInfC{$q \cdot q$ is well formed}
\end{prooftree}

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$q'$ is well formed}
\AxiomC{$W(q) \cap W(q') = \varnothing$}
\TrinaryInfC{$q \cdot q$ is well formed}
\end{prooftree}

@@

  \caption{QRE Inference Rules}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended meaning:
\begin{theorem}

@@

  \item $W(q)$ and $K(q)$ are well defined regular expressions,

@@

\end{theorem}
